The galactic chemical evolution of carbon: Implications for stellar nucleosynthesis

### 2025-11-25 carbon_paper/tex/main.tex
arXiv actual submission version

% AGB yield tables
\newcommand{\fruity}{\texttt{\hyperlink{fruity}{Fruity}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{2.57}
\newcommand{\nsubgiants}{14,066}

@@

\footnotetext{In this paper, we use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the mass ratio between A and B, scaled such that $[A/B]=0$ for the Sun (see Table~\ref{tab:fiducial_mod} for solar scale.) 
}

@@

We explore four different AGB star yield tables from the literature, which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Each AGB yield table reports yields for stellar models on a grid of masses and metallicities as reported in Table~\ref{tab:agb}.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. (See Appendix~\ref{sec:nugrid_yields} for the implementation and recalculation of these yields.)
For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \ycagb(m, Z) \rightarrow \aagb\ \ycagb(m, Z).
\end{equation}
As our fiducial AGB yield, we use the \fruity\ table with $\aagb=\cfactor$, (balancing MW and gas-phase constraints; see Section~\ref{sec:results} and Section~\ref{sec:gas}).

Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models, which we leave unmodified (i.e., $\aagb = 1$) for this comparison.
Most models agree on general yield trends in mass and metallicity.
C production peaks between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly. As a result, these models predict AGB C production to decrease and become more rapid with increasing metallicity.

@@

Both HBB and TDU result in mass and metallicity-dependent C yields. 
Lower mass stars ($\lesssim 1\,\Mo$)  do not experience (strong) TDU. As a result, C yields from these stars are affected only by first dredge-up, resulting in small net C yields \citep[table 1]{karakas10}.
Above ${\sim}1\,\Mo$, TDU becomes more important, enriching the outer layers with C.
In AGB stars more massive than ${\sim}5\,\Mo$, efficient HBB turns most of the $^{12}$C into $^{14}$N.
TDU is more efficient for compact, massive cores. HBB requires a more massive AGB star to reach sufficient temperatures at the base of the convective envelope to initiate the CNO cycle. Both processes are more efficient at low metallicity due to the lower opacities, hotter temperatures, and more compact internal structures. The details of each process are ultimately set by uncertain choices in stellar evolution models.

@@

%\dbnote{also NUGRID... https://ui.adsabs.harvard.edu/abs/2012ApJ...757..132H/abstract, https://ui.adsabs.harvard.edu/abs/2016MNRAS.455.3848J/abstract}

For our comparisons later, we define the AGB yield fraction, 
\begin{equation}\label{eq:f_agb}
    \fagb := \frac{\Ycagb(\Zo)}{\Ycc(\Zo) + \Ycagb(\Zo)}.
    \end{equation}
This quantity should approximately represent the fraction of the Sun's birth C abundance originating from AGB stars. However, since the Sun was not enriched entirely by solar metallicity predecessors, the equivalence of $\fagb{}$ to the yield ratio is not exact.

@@

CCSN models predict a wide range of C yields, spanning nearly a factor of ten, even at fixed metallicity.
Fig.~\ref{fig:y_cc} shows the integrated yields of massive star models from the literature. At solar metallicity, \citepos{sukhbold+16} C yield is $\approx0.5$ dex higher than \cite[non-rotating]{LC18}, likely driven by much stronger (C-enhanced) mass loss in \citet{sukhbold+16} for stars more massive than $\gtrsim40\,\Mo$.
 The \cite{LC18} models, which include rotation, show that the induced mixing (e.g., \citealt{frischknecht+16}) can dramatically increase the magnitude and metallicity dependence of $\Ycc$. As we will later empirically show, CCSN C production needs to be strongly metallicity-dependent at $Z \approx \Zo$, which is consistent with \citepos{LC18} rapidly rotating models and to a lesser extent \citet{NKT13}.
Fig.~\ref{fig:y_cc} also shows the IMF-averaged AGB star yields from the \fruity\ model (in Fig.~\ref{fig:agb-ssp}). Especially at $Z\approx \Zo$, most CCSN models dominate AGB C production.

@@

\begin{figure}
    \centering
    \includegraphics{cc_yields.pdf}
    \caption[]{
        IMF-integrated C yields from massive stars plotted as a function of metallicity.
        The right axis provides the equivalent CCSN [C/Mg]  ratio, assuming our fiducial $y_{\rm Mg}$ yield. The black line is our fiducial massive star yield (see Eq.~\ref{eq:y_cc}).
        Yields are shown for tables from \citet[LC18, blue circles]{LC18}, \nugrid{} (P16, orange hexagons),  \citet[S16, green square and diamond]{sukhbold+16},  \citet[NKT13, red stars]{NKT13}, and 
    \citet[WW95, pink triangles]{WW95}. \citet{sukhbold+16} report yields for different black hole landscapes, while \citet{LC18} provide yields at different rotational velocities.
   The pink line denotes $\Ycagb$ from the \fruity{} AGB star model for comparison (see Fig.~\ref{fig:agb-ssp}). All models include wind yields. 
}
    \label{fig:y_cc}
\end{figure}

@@

Our models extend the \citet[hereafter \JJ]{james+21} Milky Way model. We integrate the model using the publicly available Versatile Integrator for Chemical Evolution (\VICE; \citealt{JW20}).%
    \footnote{\VICE~is available at \url{https://github.com/giganano/VICE}}
This model is described extensively in \JJ~and summarized in \citet{james+23}. Here, we provide an overview of the relevant components.

@@

We adopt the following priors for each model parameter: $\aagb \sim N(1, 1)$, $\zcc \sim N(0, 10^{-3})$, and $\yl \sim N(2\times10^{-3}, 1\times10^{-3})$, where $N(\mu, \sigma)$ is a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$. 
We run the MCMC using Turing.jl \citep{turingjl} with a No-U-turn sampler (an adaptive Hamilton Monte-Carlo algorithm, \citealt{NUTS}) with an acceptance rate of 0.65. We combine 16 independent chains with 3,000 steps each for a total of 48,000 samples.
The median parameters derived from this model provide the yield choices for the AGB yield models shown in the left panel of Fig.~\ref{fig:sims_degens}.

@@

As a final test of our model, we compare the predictions against C abundances
in the gas phase. We consider measurements in Galactic and extragalactic \Hii{} regions, Damped Lyman-$\alpha$ systems\footnote{damped Lyman-$\alpha$ systems (or DLAs) are gas clouds observed as absorption spectra in front of background quasars}, and Milky Way halo main-sequence stars (all references in the caption of Fig.~\ref{fig:gas_phase}, which we do not repeat here for brevity).
While we used Mg as our representative alpha element in our APOGEE sample, we shift focus to O in this Section because O is more readily observed in the gas phase.
C abundances are challenging to measure in \Hii{} regions because of a lack of strong recombination lines. The collisionally excited lines of C fall in the far ultraviolet without nearby reference H lines \citep[][]{skillman+20}. 
%Furthermore, CEL and RL derived abundances are systematically discrepant by \about{0.5}\,dex \citep{GR07}.
Metals may also be trapped in dust grains or in unobserved ionization states, requiring model-dependent corrections \citep[e.g.,][]{MM19}.
We will see below that, despite these challenges, a global trend in [C/O] with metallicity nevertheless emerges.
 % DLAs, on the other hand, are relatively precise because these regions are lower-ionization, have a well-defined region (column density), and often have high-quality spectra covering rest-frame UV and optical {\color{red} CITE/verify}.

@@

From the combined constraints on the [C/Mg] trend with metallicity and [$\alpha$/Fe], we have inferred possible combinations of yields that reproduce thin-disk abundance trends. In all cases, we find that $\Yct / \Ymg \approx 4.2 \pm 0.2$ at solar metallicity, with total C production increasing with increasing metallicity. We also show that low-intermediate mass AGB stars likely produce a fraction of between 0.10 and 0.40 of solar C, across GCE assumptions and AGB yield tables.
At the normalization of stellar yields recommended by \citep{david_fe},  most AGB star models predict C yields within a factor of ${\sim}2$ of our inference.  We find that the best-fitting models require between ${\sim}0.6$ and 2 times more C than predicted by the chosen AGB table.

@@

	\begin{tabular}{l l l l l l l l l}
		\hline
            name & $\aagb$ & $\ycagb$ & $\yo/10^{-4}$ & $\zcc/10^{-4}$ \\ 
            \hline
            fiducial & 2.57 & fruity & 18.7 & 7.9 \\
            $\zcc=0.0004$ & 2.57 & fruity & 18.7 & 4.0 \\
            $\zcc=0.0012$ & 2.57 & fruity & 18.7 & 11.9 \\
            $\fagb = 0$ & 0.0 & fruity & 27.1 & 7.9 \\
            $\fagb = 0.5$ & 4.36 & fruity & 13.3 & 7.9 \\
            lower $\fagb$, $\zcc$ & 0 & fruity & 27.9 & 4.1 \\
            higher $\fagb$, $\zcc$ & 4.36 & fruity & 12.1 & 12.1 \\
            \fruity{} & 1.55 & fruity & 22.3 & 7.2 \\
            \aton{} & 1.07 & aton & 26.2 &  11.3 \\
            \monash{} & 1.12 & monash & 23.7 & 10.9 \\
            \nugrid{} & 0.59 & nugrid & 22.3 & 4.5 \\
            \fruity{} shifted & 1.29 & fruity(m/0.7) & 23.3 & 4.6 \\
            mass shift 0.5 & 2.98 & fruity(m/0.5)  & 22.1 & 14.5 \\
            mass shift 0.7 & 2.34 & fruity(m/0.7) & 19.7 & 3.2 \\
            mass shift 1.5 & 2.71 & fruity(m/1.5) & 18.6 & 8.5 \\
		\hline
	\end{tabular}
\end{table}

@@

Because the yield tables included in \citep{battino+19, battino+21} are inconsistent with the \citet{ritter+18} definitions of yields, we recalculate these yields from the publicly available data on \url{https://astrohub.uvic.ca}.

>>>>

% AGB yield tables
\newcommand{\fruity}{\texttt{\hyperlink{fruity}{FRUITY}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{2.57}
\newcommand{\nsubgiants}{14,066}

@@

\footnotetext{In this paper, we use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the mass ratio between A and B, scaled such that $[A/B]=0$ for the Sun. (See Table~\ref{tab:fiducial_mod} for the solar scale).
}

@@

We explore four different AGB star yield tables from the literature, which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Each AGB yield table reports yields for stellar models on a grid of masses and metallicities as reported in Table~\ref{tab:agb}.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. (See Appendix~\ref{sec:nugrid_yields} for the implementation and recalculation of these yields.)

Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models.
Most models agree on general yield trends in mass and metallicity.
C production peaks between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly. As a result, these models predict AGB C production to decrease and become more rapid with increasing metallicity.

@@

Both HBB and TDU result in mass and metallicity-dependent C yields. 
Lower mass stars ($\lesssim 1\,\Mo$)  do not experience (strong) TDU. As a result, C yields from these stars are affected only by first dredge-up, resulting in small net C yields \citep{karakas10}.
Above ${\sim}1\,\Mo$, TDU becomes more important, enriching the outer layers with C.
In AGB stars more massive than ${\sim}5\,\Mo$, efficient HBB turns most of the $^{12}$C into $^{14}$N.
TDU is more efficient for compact, massive cores. HBB requires a more massive AGB star to reach sufficient temperatures at the base of the convective envelope to initiate the CNO cycle. Both processes are more efficient at low metallicity due to the lower opacities, hotter temperatures, and more compact internal structures. The details of each process are ultimately set by uncertain choices in stellar evolution models.

@@

For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \ycagb(m, Z) \rightarrow \aagb\ \ycagb(m, Z).
\end{equation}
We also define the AGB yield fraction, 
\begin{equation}\label{eq:f_agb}
    \fagb := \frac{\Ycagb(\Zo)}{\Ycc(\Zo) + \Ycagb(\Zo)}.
    \end{equation}
This quantity should approximately represent the fraction of the Sun's birth C abundance originating from AGB stars. However, since the Sun was not enriched entirely by solar metallicity predecessors, the equivalence of $\fagb{}$ to the yield ratio is not exact.
As our fiducial AGB yield, we use the \fruity\ table with $\aagb=\cfactor$ ($\fagb=0.3$), balancing MW and gas-phase constraints (see Sections~\ref{sec:results} and \ref{sec:gas}).

@@

CCSN models predict a wide range of C yields, spanning nearly a factor of ten, even at fixed metallicity.
Fig.~\ref{fig:y_cc} shows the integrated yields of massive star models from the literature. At solar metallicity, \citepos{sukhbold+16} C yield is $\approx0.5$ dex higher than \citepos{LC18} non-rotating models, likely driven by much stronger (C-enhanced) mass loss in \citet{sukhbold+16} for stars more massive than $\gtrsim40\,\Mo$.
 The \cite{LC18} models including rotation show that the induced mixing (e.g., \citealt{frischknecht+16}) can dramatically increase the magnitude and metallicity dependence of $\Ycc$. As we will later empirically show, CCSN C production needs to be strongly metallicity-dependent at $Z \approx \Zo$, which is consistent with \citepos{LC18} rapidly rotating models and to a lesser extent \citet{NKT13}.
Fig.~\ref{fig:y_cc} also shows the IMF-averaged AGB star yields from the \fruity\ model (from Fig.~\ref{fig:agb-ssp}). Especially at $Z\approx \Zo$, most CCSN models dominate AGB C production.

@@

\begin{figure}
    \centering
    \includegraphics{cc_yields.pdf}
    \caption[]{
        IMF-integrated C yields from massive stars plotted as a function of metallicity.
        The right axis provides the equivalent CCSN [C/Mg]  ratio, assuming our fiducial $y_{\rm Mg}$ yield. The black line is our fiducial massive star yield (see Eq.~\ref{eq:y_cc}).
        Yields are shown for tables from \citet[LC18, blue circles]{LC18}, \nugrid{} (P16, orange hexagons),  \citet[S16, green square and diamond]{sukhbold+16},  \citet[NKT13, red stars]{NKT13}, and 
    \citet[WW95, pink triangles]{WW95}. \citet{sukhbold+16} report yields for different black hole landscapes, while \citet{LC18} provide yields at different rotational velocities.
   The light blue line denotes $\Ycagb$ from the \fruity{} AGB star model for comparison (see Fig.~\ref{fig:agb-ssp}). All models include wind yields. 
}
    \label{fig:y_cc}
\end{figure}

@@

Our models extend the \citet[hereafter \JJ]{james+21} Milky Way model. We integrate the model using the publicly available Versatile Integrator for Chemical Evolution (\VICE; \citealt{JW20}).%
    \footnote{\VICE~is available at \url{https://github.com/giganano/VICE}.}
This model is described extensively in \JJ~and summarized in \citet{james+23}. Here, we provide an overview of the relevant components.

@@

We adopt the following priors for each model parameter: $\aagb \sim N(1, 1)$, $\zcc \sim N(0, 10^{-3})$, and $\yl \sim N(2\times10^{-3}, 1\times10^{-3})$, where $N(\mu, \sigma)$ is a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$. (For the doubled yields and outflows model, the priors are correspondingly doubled). 
We run the MCMC using Turing.jl \citep{turingjl} with a No-U-turn sampler (an adaptive Hamilton Monte-Carlo algorithm, \citealt{NUTS}) with an acceptance rate of 0.65. We combine 16 independent chains with 3,000 steps each for a total of 48,000 samples.
The median parameters derived from this model provide the yield choices for the AGB yield models shown in the left panel of Fig.~\ref{fig:sims_degens}.

@@

As a final test of our model, we compare the predictions against C abundances
in the gas phase. We consider measurements in Galactic and extragalactic \Hii{} regions, Damped Lyman-$\alpha$ systems,\footnote{Damped Lyman-$\alpha$ systems (or DLAs) are gas clouds observed as absorption spectra in front of background quasars.} and Milky Way halo main-sequence stars (all references in the caption of Fig.~\ref{fig:gas_phase}, which we do not repeat here for brevity).
While we used Mg as our representative alpha element in our APOGEE sample, we shift focus to O in this Section because O is more readily observed in the gas phase.
C abundances are challenging to measure in \Hii{} regions because of a lack of strong recombination lines. The collisionally excited lines of C fall in the far ultraviolet without nearby reference H lines \citep[][]{skillman+20}. 
%Furthermore, CEL and RL derived abundances are systematically discrepant by \about{0.5}\,dex \citep{GR07}.
Metals may also be trapped in dust grains or in unobserved ionization states, requiring model-dependent corrections \citep[e.g.,][]{MM19}.
We will see below that, despite these challenges, a global trend in [C/O] with metallicity nevertheless emerges.
 % DLAs, on the other hand, are relatively precise because these regions are lower-ionization, have a well-defined region (column density), and often have high-quality spectra covering rest-frame UV and optical {\color{red} CITE/verify}.

@@

From the combined constraints on the [C/Mg] trend with [Mg/H] and [Mg/Fe], we have inferred possible combinations of yields that reproduce thin-disk abundance trends. In all cases, we find that $\Yct / \Ymg \approx 4.2 \pm 0.2$ at solar metallicity, with total C production increasing with increasing metallicity. We also show that low-intermediate mass AGB stars likely produce a fraction of between 0.10 and 0.40 of solar C, across GCE assumptions and AGB yield tables.
At the normalization of stellar yields recommended by \citep{david_fe},  most AGB star models predict C yields within a factor of ${\sim}2$ of our inference.  We find that the best-fitting models require between ${\sim}0.6$ and 2 times more C than predicted by the chosen AGB table.

@@

	\begin{tabular}{l l l l l l l l l}
		\hline
            name & $\aagb$ & $\ycagb$ & $\yo/10^{-4}$ & $\zcc/10^{-4}$ \\ 
            \hline
            fiducial & 2.57 & FRUITY & 18.7 & 7.9 \\
            $\zcc=0.0004$ & 2.57 & FRUITY & 18.7 & 4.0 \\
            $\zcc=0.0012$ & 2.57 & FRUITY & 18.7 & 11.9 \\
            $\fagb = 0$ & 0.0 & FRUITY & 27.1 & 7.9 \\
            $\fagb = 0.5$ & 4.36 & FRUITY & 13.3 & 7.9 \\
            lower $\fagb$, $\zcc$ & 0 & FRUITY & 27.9 & 4.1 \\
            higher $\fagb$, $\zcc$ & 4.36 & FRUITY & 12.1 & 12.1 \\
            \fruity{} & 1.55 & FRUITY & 22.3 & 7.2 \\
            \aton{} & 1.07 & ATON & 26.2 &  11.3 \\
            \monash{} & 1.12 & Monash & 23.7 & 10.9 \\
            \nugrid{} & 0.59 & NuGrid & 22.3 & 4.5 \\
            \fruity{} shifted & 1.29 & FRUITY($M/0.7$) & 23.3 & 4.6 \\
            mass shift 0.5 & 2.98 & FRUITY($M/0.5$)  & 22.1 & 14.5 \\
            mass shift 0.7 & 2.34 & FRUITY($M/0.7$) & 19.7 & 3.2 \\
            mass shift 1.5 & 2.71 & FRUITY($M/1.5$) & 18.6 & 8.5 \\
		\hline
	\end{tabular}
\end{table}

@@

Because the yield tables included in \citet{battino+19, battino+21} are inconsistent with the \citet{ritter+18} definitions of yields, we recalculate these yields from the publicly available data on \url{https://astrohub.uvic.ca}.
